Estimation of the effect of a continuous treatment variable

### 2025-10-26 Writing/manuscript_CF_continuous.qmd
major update on writing plus introduction

---
title: "Estimation of the effect of a continuous treatment variable"
format:
  html: default
execute:
  echo: false
bibliography: references.bib
---

```{r}
#| include: false
library(data.table)
library(dplyr)
library(flextable)
library(officer)
```

@@

+ Background
+ Why we need the methods we are proposing? Motivation.
+ Summary of findings

@@

We run Monte Carlo simulation to test the performance of ...

@@

<span style = "color: blue;"> Use math here to explain orthogonalization. This is where you explain what SO and DO in general. The sections below add specifics to the individual approaches we take. You also need to provide contexts (past studies). </span>

@@

In the on farm precision experiment (OFPE) setting, the ideas behind DML have shown promise for causal effect estimation. @kakimoto2022causal applied this framework to estimate heterogeneous treatment effects of nitrogen on yield, but their implementation was limited to a single-year experiment from one field and treated heterogeneity in a discrete way. They defined multiple “treatments” as increases from the lowest nitrogen rate to each higher rate, which is natural when there are only five or six fixed target rates per field. When data are pooled across fields with differing targets, however, applied nitrogen rates become effectively continuous, and the standard residualization of the treatment in DML does not extend cleanly to that setting.

@@

For each simulation round, one field is selected as the test field. The remaining fields are used to construct the training set in three scenarios: (1) one-field training, (2) five-field training, and (3) ten-field training. These scenarios allow us to assess how model performance changes with the size and diversity of training data. Within each round, the model is trained on the designated number of training fields and then used to estimate the yield response function and the site-specific EONR for the test field.

@@

@@

We compare six estimators—CF, SO_ANN, DO_ANN, NO_ANN, SO_RF, and NO_RF—over 500 Monte-Carlo replications and three training sizes (1, 5, and 10 fields). Each replication contains 1,440 evaluation units. Accuracy is assessed by the RMSE of the estimated economically optimal nitrogen rate (EONR) with respect to the true (simulated) EONR. Figures 1–4 summarize errors; Figure 5 panels illustrate representative yield–N response functions that help interpret the error patterns.

```{r setup, include=FALSE}

@@

# load data
rmse_tbl_raw <- readRDS(here::here("results", "rmse_tbl_raw.rds"))
```

@@

```{r}
#| label: fig-rmse-hist
#| fig-cap: "RMSE histograms by model and number of fields."
#| fig-width: 12
#| fig-height: 7

# model names: row labels in the graph
model_to_row <- c(
  "Simple_ANN" = "No_ANN",
  "RANN"       = "DO_ANN",
  "NO_RF"      = "NO_RF",
  "HalfRANN"   = "SO_ANN",
  "HalfRRF"    = "SO_RF",
  "CF"         = "CF"
)

rmse_plot_df <- rmse_tbl_raw %>%
  dplyr::filter(ok_rows, Model %in% names(model_to_row)) %>%
  dplyr::mutate(
    row_lab = dplyr::recode(Model, !!!model_to_row),
    col_lab = factor(`#field`,
                     levels = c(1, 5, 10),
                     labels = c("Number of fields: 1",
                                "Number of fields: 5",
                                "Number of fields: 10"))
  )

# consistent axes
xmin <- 0
nice_ceiling <- function(x, base = 5) base * ceiling(x / base)
xmax <- nice_ceiling(max(rmse_plot_df$RMSE, na.rm = TRUE), base = 5)
bins <- 25
binwidth <- (xmax - xmin) / bins

p <- ggplot(rmse_plot_df, aes(x = RMSE)) +
  geom_histogram(binwidth = binwidth, boundary = 0, closed = "left") +
  facet_grid(row_lab ~ col_lab, scales = "fixed") +
  scale_x_continuous(limits = c(xmin, xmax),
                     breaks = seq(xmin, xmax, length.out = 6),
                     expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "RMSE by iteration", y = "Count") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 14)
  )

p
```

Mean RMSE by model and training size (@fig-avg-performance and @tbl-mean-rmse-heatmap) confirms these patterns. CF is consistently low (≈15 across sizes). SO_ANN improves monotonically as data increase and attains the best mean at 10 fields (≈11). NO_ANN is unusable with a single field (mean ≈82) but becomes competitive by 5–10 fields (≈14 and ≈13). SO_RF yields moderate means (≈22 at 10 fields), while NO_RF remains highest among the non-ANN baselines (≈36–40). DO_ANN is uniformly poor (≈90–93). The heatmap (Fig. 3) prints the means in each cell and emphasizes the strong gains from orthogonalization (SO_ANN and CF) relative to their non-orthogonal counterparts.

@@

rmse_df <- rmse_tbl_raw %>%
  dplyr::group_by(Model, `#field`, sim) %>%  
  dplyr::summarise(RMSE = dplyr::first(RMSE), .groups = "drop") %>%
  # Rename models
  dplyr::mutate(
    Model = dplyr::recode(
      Model,
      "Simple_ANN" = "NO_ANN",
      "HalfRANN"   = "SO_ANN",
      "RANN"       = "DO_ANN",
      "HalfRRF"    = "SO_RF",
      "CF"         = "CF",
      "NO_RF"      = "NO_RF",
      .default     = Model
    ),
    field_f = factor(`#field`, levels = c(1, 5, 10),
                     labels = c("1 field", "5 fields", "10 fields"))
  )

# 3) summary stats 
rmse_summary <- rmse_df %>%
  dplyr::group_by(Model, field_f) %>%
  dplyr::summarise(
    n      = dplyr::n(),
    mean   = mean(RMSE, na.rm = TRUE),
    median = median(RMSE, na.rm = TRUE),
    sd     = sd(RMSE, na.rm = TRUE),
    se     = sd / sqrt(n),
    lower  = mean + qt(0.025, df = n - 1) * se,
    upper  = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

# 
p_mean <- ggplot(rmse_summary, aes(x = field_f, y = mean, group = Model, color = Model)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 3) +
  labs(x = "# fields", y = "Mean RMSE", color = "Model") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 13),
    strip.text = element_text(face = "bold")
  )

@@

```{r}
#| label: tbl-mean-rmse-heatmap
#| tbl-cap: "Mean RMSE by model and number of fields."
#| tbl-cap-location: bottom
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(gt)
library(scales)

# Range for coloring
fill_min <- floor(min(rmse_summary$mean, na.rm = TRUE))
fill_max <- ceiling(max(rmse_summary$mean, na.rm = TRUE))

# Wide table of means
rmse_table <- rmse_summary %>%
  select(Model, field_f, mean) %>%
  mutate(mean = round(mean, 1)) %>%
  pivot_wider(names_from = field_f, values_from = mean) %>%
  arrange(Model)

# Light → mid-blue palette
pal <- col_numeric(
  palette = c("#f7fbff", "#6baed6"),
  domain  = c(fill_min, fill_max)
)

# HTML colorbar legend to place under the table
legend_html <- sprintf(
  '<div style="display:flex;align-items:center;gap:8px;margin-top:6px;">
     <span style="font-size:12px;color:#444;">Low (%.1f)</span>
     <div style="flex:1;height:10px;background:linear-gradient(to right,#f7fbff,#6baed6);
                 border:1px solid #ddd;border-radius:3px;"></div>
     <span style="font-size:12px;color:#444;">High (%.1f)</span>
   </div>', fill_min, fill_max)

rmse_table %>%
  gt(rowname_col = "Model") %>%
  data_color(
    columns = c(`1 field`, `5 fields`, `10 fields`),
    colors  = pal
  ) %>%
  fmt_number(columns = c(`1 field`, `5 fields`, `10 fields`), decimals = 1) %>%
  cols_label(
    `1 field`  = "1 field",
    `5 fields` = "5 fields",
    `10 fields`= "10 fields"
  ) %>%
  tab_options(
    table.font.size = px(13),
    data_row.padding = px(6),
    column_labels.font.weight = "bold"
  ) %>%
  tab_source_note(source_note = gt::html(legend_html))
```

@@

>>>>

---
title: "Estimation of the effect of a continuous treatment variable"
format:
  docx:
    toc: true

execute:
  echo: false
bibliography: references.bib
---

```{r}
#| include: false
library(data.table)
library(dplyr)
library(flextable)
library(officer)
library(data.table)
library(tidyverse)
library(officedown)
library(officer)
library(flextable)
library(stringr)
library(sf)
library(lfe)
library(modelsummary)
library(patchwork)
library(gridExtra)
library(kableExtra)
library(measurements)
```

@@

Efficient nitrogen (N) management remains one of the most critical challenges in modern agricultural production. Nitrogen is essential for achieving high crop yields, yet excessive or poorly timed applications can lead to economic losses and environmental harm. The principle of applying the right amount of nitrogen at the right place and time requires understanding the spatial and temporal variability in crop nitrogen needs. This variability arises from complex interactions among soil nitrogen availability, soil and field characteristics, and weather conditions. Accurately identifying the optimal nitrogen rate for each site and season is therefore fundamental to improving both profitability and sustainability in crop production systems.

On-farm field experiments have emerged as a powerful approach to generate the data necessary for site-specific nitrogen management. Through on-farm agronomic trials, researchers can directly observe yield responses to varying nitrogen rates under real-world field conditions. Analyzing these data enables estimation of yield response functions and identification of site-specific economically optimal nitrogen rates. Recent advances in data analytics—particularly the widespread adoption of machine learning (ML) methods—have expanded the potential of on-farm experimentation. Two major methodological directions have emerged: (1) yield prediction–based methods, which recommend site-specific input rates by predicting yield as a function of nitrogen and environmental covariates (@krause2020random; @barbosa2020modeling; @barbosa2020risk; @gardner2021economic), and (2) causal machine learning (causal ML) methods (@kakimoto2022causal), which estimate the underlying causal relationship between nitrogen application and yield response.

In the OFPE setting, causal machine learning methods have shown to be promising to recover the full yield response curve. @kakimoto2022causal showed that causal forest can estimate heterogeneous treatment effects of an input on yield more accurately than prediction-oriented machined approaches like random forest, boosted forest, and convolutional neural network. However, their application of CF is only limited to a single-year OFPE experiment from a single field. Causal forest was developped primarily to estimate
the heterogeneous treatment effects of a dicrete treatment. In OFPE, there are typically five
or six distinctive target experimental rates (see the first panel of @fig-distribution-N). In @kakimoto2022causal, CF estimates the effect of multiple treatments defined as the increase in nitrogen rate from the lowest rate to the other rates (so n − 1 tretments in n-rate trials). However,
once data from multiple fields with different target rates are combined, applied nitrogen rates
become rather continuous as seen in the second panel of @fig-distribution-N. Existing approaches—such as direct applications of random forests or neural networks and CF—often tend to fail to capture heterogeneous treatment effects of a continuous variable (@kakimoto2022causal).

```{r n-histogram, include = FALSE}
# sim_data <-
#   readRDS(here::here("data/raw_sim_data.rds")) %>%
#    data.table::data.table()
# 
# reg_data_all <-
#   sim_data[, .(design_name, reg_data)] %>%
#   unnest(reg_data) %>%
#   data.table::data.table() %>%
#   .[, design := case_when(
#     design_name == "Latin Square Fixed 5" ~ "ls5",
#     design_name == "Latin Square Fixed 6" ~ "ls6"
#   )]
# 
# data_guide <-
#   lapply(
#     1:nrow(reg_data_all),
#     function(x) {
#       temp_data <- reg_data_all[x, ]
#       temp_sim <- temp_data[, sim]
#       temp_design <- temp_data[, design]
# 
#       data_file_name <- paste0("data/sim_data_", temp_sim, "_design_", temp_design, ".rds")
# 
#       saveRDS(
#         temp_data,
#         here::here(data_file_name)
#       )
# 
#       return(
#         data.table::data.table(
#           sim = temp_sim,
#           design = temp_design,
#           file_name = data_file_name
#         )
#       )
#     }
#   ) %>%
#   data.table::rbindlist()
# 
# saveRDS(data_guide, here::here("data/data_guide.rds"))

# sim_des_ls <- data_guide[, sim_des]
# train_site_num <- 20
# 
# set.seed(87924354)
# 
# test_train <-
#   data_guide %>%
#   rowwise() %>%
#   mutate(train_sim = list(
#     sample(
#       sim_des_ls[sim_des_ls != sim_des],
#       train_site_num
#     )
#   )) %>%
#   data.table::data.table()
# 
# saveRDS(test_train, here::here("data/test_train_ids.rds"))

#####

# data_guide <-
#   readRDS(here::here("data/data_guide.rds")) %>%
#   .[, sim_des := paste0(sim, "_", design)]
# 
# test_train <- readRDS(here::here("data/test_train_ids.rds"))
# 
# temp_data <- test_train[8, ]
# 
# test_sim_des <- temp_data$sim_des
# 
# #--- get the sim_des ids for training data ---#
# train_sim_des <- temp_data$train_sim[[1]]
# 
# #--- train data ---#
# train_data <-
#   data_guide[sim_des %in% train_sim_des, file_name] %>%
#   lapply(., function(x) readRDS(here::here(x))) %>%
#   data.table::rbindlist() %>%
#   .[, data] %>%
#   data.table::rbindlist() %>%
#   .[, type := "Multiple Fields"]
# 
# #--- test data ---#
# test_data <-
#   readRDS(here::here(temp_data$file_name))$data[[1]] %>%
#   .[, type := "Single Field"]
# 
# data_illust <-
#   rbind(test_data, train_data) %>%
#   .[, type := factor(type, levels = c("Single Field", "Multiple Fields"))]
# 
# (
#   g_N_hist_illust <-
#     ggplot(data_illust) +
#     geom_histogram(
#       aes(x = N),
#       bins = 50,
#       fill = "gray",
#       color = "blue"
#     ) +
#     facet_grid(type ~ ., scales = "free_y") +
#     xlab("Nitrogen Application Rate (lb/acre)") +
#     ylab("Count")
# )

# ggsave(
#   g_N_hist_illust,
#   file = here::here("writing/Figures/g_N_hist_single_multiple.pdf"),
#   width = 6.5,
#   height = 4
# )

```{r fig-distribution-N, fig.cap="Distribution of nitrogen rates from a single OFPE and multiple OFPEs"}
knitr::include_graphics(here::here("writing/Figures/g_N_hist_single_multiple.pdf"))
```
As multi-year, multi-field OFPE datasets accumulate, pooling data across fields has become both feasible and desirable—it can improve estimation of yield response functions and deliver stronger agronomic and economic insights. It creates a need for methods that handle continuous treatments and it becomes especially important to adapt causal methods to continuous treatments without losing their statistical strengths over other machine-learning approaches.

The key objective of this article is to identify a causal machine learning approach capable of estimating heterogeneous treatment effects for yield response function estimation using data from on-farm trials, where the treatment variable—nitrogen rate—is continuous and randomized. We investigate two complementary approaches: a Single-Orthogonalized Neural Network (SO-ANN), designed to flexibly capture nonlinear treatment–response relationships, and an extended Causal Forest (CF) that incorporates smoothing splines to model continuous treatments. Both methods aim to preserve the strengths of causal ML—nonparametric modeling of complex interactions without restrictive assumptions. Their performance is evaluated through Monte Carlo simulations that mimic typical OFPE conditions and crop yield responses to nitrogen.

We find that the SO-ANN approach outperforms the non-orthogonalized model, which represents the common practice in yield response function estimation. In contrast, the Double-Orthogonalized approach performs poorly, and its accuracy does not improve even with larger sample sizes, consistent with theoretical expectations. Our extended CF method successfully captures heterogeneous and nonlinear effects of a continuous treatment and accurately estimates site-specific profit-maximizing nitrogen rates with minimal bias underscoring the promise of this approach. Although the methodological extension proposed here is primarily motivated by the needs of OFPE, it holds potential for a wide range of applications in applied economics where continuous treatments and heterogeneous effects are of interest.

@@

@@

@@

@kakimoto2022causal applied this framework to estimate heterogeneous treatment effects of nitrogen on yield and treated heterogeneity in a discrete way. However, when data are pooled across fields and applied nitrogen rates become continuous, the standard residualization of the treatment in DML does not extend cleanly to that setting.

@@

For each simulation round, one field is selected as the test field. The remaining fields are used to construct the training set in three scenarios: (1) one-field training, (2) three-field training, (3) five-field training, (4) ten-field training and (5) twenty-field training. These scenarios allow us to assess how model performance changes with the size and diversity of training data. Within each round, the model is trained on the designated number of training fields and then used to estimate the yield response function and the site-specific EONR for the test field.

@@

First, a 92.17-acre rectangular field is created, which consists of $384$ of `r measurements::conv_unit(18.3, "m", "ft") %>% round(digits = 0)` feet $\times$ `r measurements::conv_unit(73.2, "m", "ft") %>% round(digits = 0)` feet plots. each of which will be assigned an N fertilizer application rate. Each plot is made up of four (4-rows $\times$ 1-column) `r measurements::conv_unit(18.3, "m", "ft") %>% round(digits = 0)` feet $\times$ `r measurements::conv_unit(18.3, "m", "ft") %>% round(digits = 0)` feet "subplots," which are the unit of analysis used in subsequent statistical analysis. Each subplot consists of a 6-rows $\times$ 6-columns grid of thirty-six `r measurements::conv_unit(3.05, "m", "ft") %>% round(digits = 0)` feet $\times$ `r measurements::conv_unit(3.05, "m", "ft") %>% round(digits = 0)` feet "cells." 
Visual representation of the field layout is presented in @fig-field-layout. The data are generated at the cell-level first and then aggregated up to analysis-unit level.

```{r}

# /*+++++++++++++++++++++++++++++++++++
#' ## Plots
# /*+++++++++++++++++++++++++++++++++++

#library(patchwork)

# all_data <- readRDS(here::here("data/raw_sim_data.rds"))
# field_sf <- all_data$field_sf[[1]]
# 
# 
# plot_sf <-
#   field_sf %>%
#   nest_by(plot_id) %>%
#   mutate(data = list(
#     st_union(data)
#   )) %>%
#   unnest() %>%
#   st_as_sf()
# 
# plot_focus <- filter(plot_sf, plot_id == 283)
# 
# plot_focus_xy <-
#   st_centroid(plot_focus) %>%
#   st_coordinates()
# 
# bbox_field <- st_bbox(plot_sf)
# min_x_field <- bbox_field["xmin"]
# max_x_field <- bbox_field["xmax"]
# center_x_field <- mean(c(min_x_field, max_x_field))
# min_y_field <- bbox_field["ymin"]
# max_y_field <- bbox_field["ymax"]
# center_y_field <- mean(c(min_y_field, max_y_field))
# 
# g_field <-
#   ggplot() +
#   geom_sf(data = plot_sf, linewidth = 1, fill = NA) +
#   geom_sf(data = plot_focus, fill = "red", alpha = 0.5) +
#   #--- horizontal width ---#
#   geom_segment(
#     aes(x = center_x_field - 72, xend = min_x_field, y = max_y_field + 18, yend = max_y_field + 18),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   geom_segment(
#     aes(x = center_x_field + 72, xend = max_x_field, y = max_y_field + 18, yend = max_y_field + 18),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   annotate(
#     "text",
#     x = center_x_field,
#     y = max_y_field + 18,
#     label = "2,835 feet"
#   ) +
#   #--- vertical length ---#
#   geom_segment(
#     aes(x = max_x_field + 18, xend = max_x_field + 18, y = center_y_field + 60, yend = max_y_field),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   geom_segment(
#     aes(x = max_x_field + 18, xend = max_x_field + 18, y = center_y_field - 60, yend = min_y_field),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   annotate(
#     "text",
#     x = max_x_field + 18,
#     y = center_y_field,
#     label = "1,417 feet",
#     angle = "270"
#   ) +
#   #--- expansion to the focus plot ---#
#   geom_segment(
#     aes(x = plot_focus_xy[1, "X"] - 36, xend = min_x_field, y = plot_focus_xy[1, "Y"] - 9, yend = plot_focus_xy[1, "Y"] - 120),
#     color = "red"
#   ) +
#   geom_segment(
#     aes(x = plot_focus_xy[1, "X"] + 36, xend = max_x_field, y = plot_focus_xy[1, "Y"] - 9, yend = plot_focus_xy[1, "Y"] - 120),
#     color = "red"
#   ) +
#   annotate(
#     "text",
#     y = plot_focus_xy[1, "Y"],
#     x = plot_focus_xy[1, "X"],
#     label = "Plot",
#     size = 3
#   ) +
#   theme_void()
# 
# # /*+++++++++++++++++++++++++++++++++++
# #' ## inside a plot
# # /*+++++++++++++++++++++++++++++++++++
# plot_sf_focus <- filter(field_sf, plot_id == 1)
# 
# bbox_plot <- st_bbox(plot_sf_focus)
# min_x_plot <- bbox_plot["xmin"]
# max_x_plot <- bbox_plot["xmax"]
# center_x_plot <- mean(c(min_x_plot, max_x_plot))
# min_y_plot <- bbox_plot["ymin"]
# max_y_plot <- bbox_plot["ymax"]
# center_y_plot <- mean(c(min_y_plot, max_y_plot))
# 
# subplot_sf <-
#   plot_sf_focus %>%
#   nest_by(aunit_id, buffer) %>%
#   mutate(data = list(
#     st_union(data)
#   )) %>%
#   unnest() %>%
#   st_as_sf() %>%
#   mutate(label = ifelse(
#     buffer == 1,
#     "buffer",
#     paste0("subplot-", aunit_id - 1)
#   )) %>%
#   mutate(buf_or_not = ifelse(label == "buffer", "buffer", "subplot"))
# 
# subplot_text_sf <- st_centroid(subplot_sf)
# 
# cell_sf <- plot_sf_focus[7, ]
# 
# bbox_cell <- st_bbox(cell_sf)
# min_x_cell <- bbox_cell["xmin"]
# max_x_cell <- bbox_cell["xmax"]
# center_x_cell <- mean(c(min_x_cell, max_x_cell))
# min_y_cell <- bbox_cell["ymin"]
# max_y_cell <- bbox_cell["ymax"]
# center_y_cell <- mean(c(min_y_cell, max_y_cell))
# 
# g_inside_plot <-
#   ggplot() +
#   geom_sf(data = plot_sf_focus, size = 0.2, fill = NA) +
#   #--- horizontal width ---#
#   geom_segment(
#     aes(x = center_x_plot - 6, xend = min_x_plot, y = min_y_plot - 3, yend = min_y_plot - 3),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   geom_segment(
#     aes(x = center_x_plot + 6, xend = max_x_plot, y = min_y_plot - 3, yend = min_y_plot - 3),
#     arrow = arrow(length = unit(0.3, "cm"))
#   ) +
#   annotate(
#     "text",
#     x = center_x_plot,
#     y = min_y_plot - 3,
#     label = "236 feet"
#   ) +
#   geom_sf(data = filter(subplot_sf, buffer == 1), fill = "yellow", alpha = 0.4) +
#   geom_sf(data = subplot_sf, size = 1.2, fill = NA, color = "blue") +
#   geom_sf_text(data = subplot_text_sf, aes(label = label), size = 4, color = "blue") +
#   geom_sf(data = cell_sf, fill = "green", alpha = 0.3) +
#   #--- horizontal width of the cell ---#
#   geom_segment(
#     aes(x = center_x_cell, xend = min_x_cell, y = max_y_cell + 1, yend = max_y_cell + 1),
#     arrow = arrow(length = unit(0.2, "cm"))
#   ) +
#   geom_segment(
#     aes(x = center_x_cell, xend = max_x_cell, y = max_y_cell + 1, yend = max_y_cell + 1),
#     arrow = arrow(length = unit(0.2, "cm"))
#   ) +
#   annotate(
#     "text",
#     x = center_x_cell,
#     y = max_y_plot + 3,
#     label = "19.7 feet"
#   ) +
#   annotate("text", x = 39, y = 429, label = "cell") +
#   scale_fill_discrete(name = "") +
#   theme_void() +
#   theme(
#     legend.position = "bottom",
#     legend.text = element_text(size = 12)
#   )
# 
# g_layout <- g_field / g_inside_plot
# 
# saveRDS(g_layout, here::here("writing/Figures/g_layout.rds"))
# 
# ggsave(
#   g_layout,
#   file = here::here("writing/Figures/g_layout.pdf"),
#   width = 6.5,
#   height = 6
# )

```{r fig-field-layout, fig.cap = "Visualization of the field layout and components"}
knitr::include_graphics(here::here("Writing/Figures/g_layout.pdf"))
```
Cell-level yields are generated following the quadratic-plateau functional form as follows:

```{=latex}
\begin{align} 
y_{i,j} = 
  \begin{cases}
  \alpha_{i,j} + \beta_{i,j} N + \gamma_{i,j} N^2 + \varepsilon_{i,j}, & N < \tau_{i,j} \\
  \eta_{i,j} + \varepsilon_{i,j}, & N \geq \tau_{i,j}
  \end{cases}
\label{eq:yield-gen-sim}
\end{align}
```

In this formulation, yield increases as $N_{i,j}$ increase until $N_{i,j}$ reaches $\tau_{i,j}$, at which yield hits the plateau, $\eta_{i,j}$. The yield level at $N_{i,j}= 0$ is $\alpha_{i,j}$. The rate at which yield increases with respect to $N_{i,j}$ is governed by $\beta_{i,j}$ and $\gamma_{i,j}$. For a given cell, the quadratic-plateu function has five parameters, but the degree of freedom is only three. That is, once any three of the five parameters are determined, then the rest of the parameters are automatically determined. Specifically, we first generated $\alpha_{i,j}$, $\eta_{i,j}$, and $\tau_{i,j}$ in a spatially correlated manner following the Gaussian bluh bluh process (cite), and then $\beta_{i,j}$ and $\gamma_{i,j}$ were derived. The erorr term ($\varepsilon_{i,t}$) is generated using the same spatial process. This is repeated independenty 500 times to create 500 fields of the same size but different spatial distribution and values of the three parameters.

Trial designs vary field by field based on the value of $\tau_{i,j}$. Specifically, the lowest nitrogen rate is set at the minimum of $\tau_{i,j}$ of the field minus 50, and the highest rate is set at the maximum of $\tau_{i,j}$ of the field plus 30. We then create a sequence of equi-distance 5-level nitrogen rates. Once the unique nitrogen levels are determined, a Latin square design is used as a trial design (see  @fig-trial-design) for an example trial design). This ensures that nitrogen rate is orthogonal to any variable including the error term, avoiding endogeneity problem as done in OFPE in real world. Now, yield values at the cell level can be calculated by plugging in the value of $N_{i,j}$, $\alpha_{i,j}$, $\eta_{i,j}$, $\tau_{i,j}$, and $\varepsilon_{i,j}$ into Equation~\ref{eq:yield-gen-sim}.

```{r}
# trial_design <-
#   all_data$trial_design[[1]] %>%
#   .[sim == 1, ]
# 
# field <- data.table(field_sf)
# 
# td_sf <-
#   trial_design[field, on = c("block_id", "plot_in_block_id")] %>%
#   st_as_sf()
# 
# g_td <-
#   ggplot(td_sf) +
#   geom_sf(aes(fill = factor(N_tgt)), color = NA) +
#   scale_fill_viridis_d(name = "Nitrogen Rate (lb/acre)") +
#   theme_void() +
#   theme(legend.position = "bottom")
# 
# saveRDS(g_td, here::here("writing/Figures/g_td.rds"))
```

```{r fig-trial-design, fig.cap = "Example trial design"}
readRDS(here::here("writing/Figures/g_td.rds"))
```

@@

We compare six estimators—CF, SO_ANN, DO_ANN, NO_ANN, SO_RF, and NO_RF—over 500 Monte-Carlo replications and five training sizes (1, 3, 5, 10 and 20 fields). Each replication contains 1,440 evaluation units. Accuracy is assessed by the RMSE of the estimated economically optimal nitrogen rate (EONR) with respect to the true (simulated) EONR. Figures 1–4 summarize errors; Figure 5 panels illustrate representative yield–N response functions that help interpret the error patterns.

```{r sim-setup, include=FALSE}

@@

# load data
rmse_tbl_raw <- readRDS(here::here("results", "rmse_tbl_raw.rds"))

@@

```{r}
#| label: fig-rmse-hist
#| fig-cap: "RMSE histograms by model and number of fields."
#| fig-width: 12
#| fig-height: 7
#| fig-show: hold

library(dplyr)
library(ggplot2)
library(stringr)

model_names <- function(models) {
  map <- c(
    "Simple_ANN" = "No_ANN",
    "RANN"       = "DO_ANN",
    "HALFRANN"   = "SO_ANN",
    "HALFRRF"    = "SO_RF",
    "NO_RF"      = "NO_RF",
    "CF"         = "CF"
  )
  models_upper <- toupper(models)
  out <- map[models_upper]
  ifelse(is.na(out), models, out)
}

rmse_plot_df <- rmse_tbl_raw %>%
  filter(ok_rows) %>%
  mutate(
    row_lab = model_names(Model),
    col_lab = factor(`#field`,
      levels = sort(unique(`#field`)),
      labels = paste0("Number of fields: ", sort(unique(`#field`)))
    )
  )

# prevent table output from creating phantom figure
invisible(rmse_plot_df %>% count(row_lab, col_lab))

xmin <- 0
nice_ceiling <- function(x, base = 5) base * ceiling(x / base)
xmax <- nice_ceiling(max(rmse_plot_df$RMSE, na.rm = TRUE), base = 5)
bins <- 25
binwidth <- (xmax - xmin) / bins

p <- ggplot(rmse_plot_df, aes(x = RMSE)) +
  geom_histogram(binwidth = binwidth, boundary = 0, closed = "left", fill = "skyblue3") +
  facet_grid(row_lab ~ col_lab, scales = "fixed") +
  scale_x_continuous(limits = c(xmin, xmax), breaks = seq(xmin, xmax, length.out = 6), expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    #title = "Distribution of RMSE across models and field counts",
    x = "RMSE by iteration",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 13),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

print(p)

Mean RMSE by model and training size (@fig-avg-performance) confirms these patterns. CF is consistently low (≈15 across sizes). SO_ANN improves monotonically as data increase and attains the best mean at 10 fields (≈11). NO_ANN is unusable with a single field (mean ≈82) but becomes competitive by 5–10 fields (≈14 and ≈13). SO_RF yields moderate means (≈22 at 10 fields), while NO_RF remains highest among the non-ANN baselines (≈36–40). DO_ANN is uniformly poor (≈90–93). The heatmap (Fig. 3) prints the means in each cell and emphasizes the strong gains from orthogonalization (SO_ANN and CF) relative to their non-orthogonal counterparts.

@@

rmse_df <- rmse_tbl_raw %>%
  dplyr::group_by(Model, `#field`, sim) %>%  
  dplyr::summarise(RMSE = dplyr::first(RMSE), .groups = "drop") %>%
  dplyr::mutate(
    # Field labels (auto from unique values)
    field_f = factor(
      `#field`,
      levels = sort(unique(`#field`)),
      labels = paste0(sort(unique(`#field`)),
                      ifelse(sort(unique(`#field`)) == 1, " field", " fields"))
    )
  )

# 3) summary stats 
rmse_summary <- rmse_df %>%
  dplyr::group_by(Model, field_f) %>%
  dplyr::summarise(
    n      = dplyr::n(),
    mean   = mean(RMSE, na.rm = TRUE),
    median = median(RMSE, na.rm = TRUE),
    sd     = sd(RMSE, na.rm = TRUE),
    se     = sd / sqrt(n),
    lower  = mean + qt(0.025, df = n - 1) * se,
    upper  = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )
# 
p_mean <- ggplot(rmse_summary, aes(x = field_f, y = mean, group = Model, color = Model)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 3) +
  labs(x = "# fields", y = "Mean RMSE", color = "Model") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 13),
    strip.text = element_text(face = "bold")
  )

@@

```{r}
# #| label: tbl-mean-rmse-heatmap
# #| tbl-cap: "Mean RMSE by model and number of fields."
# #| tbl-cap-location: bottom
# #| echo: false
# #| message: false
# #| warning: false
# 
# library(dplyr)
# library(tidyr)
# library(gt)
# library(scales)
# 
# # Range for coloring
# fill_min <- floor(min(rmse_summary$mean, na.rm = TRUE))
# fill_max <- ceiling(max(rmse_summary$mean, na.rm = TRUE))
# 
# # Wide table of means
# rmse_table <- rmse_summary %>%
#   select(Model, field_f, mean) %>%
#   mutate(mean = round(mean, 1)) %>%
#   pivot_wider(names_from = field_f, values_from = mean) %>%
#   arrange(Model)
# 
# # Light → mid-blue palette
# pal <- col_numeric(
#   palette = c("#f7fbff", "#6baed6"),
#   domain  = c(fill_min, fill_max)
# )
# 
# # HTML colorbar legend to place under the table
# legend_html <- sprintf(
#   '<div style="display:flex;align-items:center;gap:8px;margin-top:6px;">
#      <span style="font-size:12px;color:#444;">Low (%.1f)</span>
#      <div style="flex:1;height:10px;background:linear-gradient(to right,#f7fbff,#6baed6);
#                  border:1px solid #ddd;border-radius:3px;"></div>
#      <span style="font-size:12px;color:#444;">High (%.1f)</span>
#    </div>', fill_min, fill_max)
# 
# rmse_table %>%
#   gt(rowname_col = "Model") %>%
#   data_color(
#     columns = c(`1 field`, `5 fields`, `10 fields`),
#     colors  = pal
#   ) %>%
#   fmt_number(columns = c(`1 field`, `5 fields`, `10 fields`), decimals = 1) %>%
#   cols_label(
#     `1 field`  = "1 field",
#     `5 fields` = "5 fields",
#     `10 fields`= "10 fields"
#   ) %>%
#   tab_options(
#     table.font.size = px(13),
#     data_row.padding = px(6),
#     column_labels.font.weight = "bold"
#   ) %>%
#   tab_source_note(source_note = gt::html(legend_html))
```

@@

```
```{r}

#| label: tbl-best_model_by_field_count
#| tbl-cap: "Best-performing model by field count."
#| tbl-cap-location: bottom
#| echo: false
#| message: false
#| warning: false

library(gt)
library(dplyr)
library(stringr)

best_models <- rmse_summary %>%
  mutate(field_n = as.numeric(str_extract(field_f, "\\d+"))) %>%
  group_by(field_n) %>%
  slice_min(order_by = mean, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(mean, field_n) %>%
  select(Model, field_f, mean)

best_models %>%
  gt() %>%
  tab_header(
    title = "Best models by #field",
    subtitle = "Ordered by lowest mean RMSE"
  ) %>%
  fmt_number(columns = c(mean), decimals = 3) %>%
  cols_label(
    field_f = "# fields",
    mean = "mean RMSE"
  ) %>%
  tab_options(table.font.size = "small")
